---
title: "The legislative network of the 116th Congress (2019-2020)"
author: "Tanner Gildea"
date: "2021-01-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(out.width = "100%")

library(tidyverse)
library(igraph)
library(ggplot2)
library(ggraph)
library(rvest)
library(httr)
library(data.table)
library(visNetwork)
library(circlize)
library(networkD3)
library(ggthemes)
library(gt)
library(reactable)

# download bills passed by at least one chamber and those that became law in 116th Congress

bills_signed <- read_csv("~/Desktop/my_projects/personal_site/content/moment/bills_signed.csv", 
                                         skip = 2) %>%
  mutate(status = "Signed into law")
dem_passed_house <- read_csv("~/Desktop/my_projects/personal_site/content/moment/dem_passed_house.csv", 
                             skip = 2)
gop_passed_house <- read_csv("~/Desktop/my_projects/personal_site/content/moment/gop_passed_house.csv", 
                              skip = 2)
passed_house <- rbind(dem_passed_house, gop_passed_house) %>%
  mutate(status = "Passed House")

senate_passed <- read_csv("~/Desktop/my_projects/personal_site/content/moment/senate_passed.csv", 
                          skip = 2) %>%
  mutate(status = "Passed Senate")

bills <- rbind(bills_signed, passed_house) %>%
  rbind(senate_passed)

bills <- bills %>% janitor::clean_names() %>%
  select(1:3,5:6,11:12,15) %>%
  mutate(cosponsor_url = paste0(url, "/cosponsors"))
```

```{r crawler, include=FALSE}
# DON'T RUN THIS CODE
### crawl/scrape bill sites with cosponsors ### 
# crawl_function <- function(x) {
#   html_content <- content(GET(x)) 
#   
#   leg_node <- html_content %>%
#     html_node('h1.legDetail') %>%
#     html_text() %>%
#     as.data.frame() %>%
#     rename(bill = 1)
#   
#   table_node <- html_content %>%
#     html_node('table.item_table') 
#   
#   if(length(table_node)>0) {
#     table_node <- table_node %>%
#       html_table() %>%
#       select(1)
#   } else {
#     cosponsor <- c('no sponsor')
#     table_node <- data.frame(cosponsor)
#   }
#   
#   left_join(leg_node, table_node, by = character())
#   }

# test <- crawl_function("https://www.congress.gov/bill/116th-congress/senate-bill/4996/cosponsors?r=4&s=1&searchResultViewType=expanded")
# test2 <- crawl_function("https://www.congress.gov/bill/116th-congress/senate-bill/5086/cosponsors?r=1&s=1")
# # no sponsors https://www.congress.gov/bill/116th-congress/senate-bill/5086/cosponsors?r=1&s=1
# multi sponsors https://www.congress.gov/bill/116th-congress/senate-bill/4996/cosponsors?r=4&s=1&searchResultViewType=expanded

# bills_signed <- bills %>% filter(status == "Signed into law")
# senate_passed <- bills %>% filter(status == "Passed Senate")
# house_passed <- bills %>% filter(status == "Passed House")
# 
# cosponsor_list_signed <- list()
# 
# for(i in 1:nrow(bills_signed)) {
#   new_table <- crawl_function(bills_signed$cosponsor_url[i])
#   cosponsor_list_signed[[length(cosponsor_list_signed) + 1]] <- new_table
# }
# 
# cosponsor_list_senate <- list()
# 
# for(i in 1:nrow(senate_passed)) {
#   new_table <- crawl_function(senate_passed$cosponsor_url[i])
#   cosponsor_list_senate[[length(cosponsor_list_senate) + 1]] <- new_table
# }
# 
# cosponsor_list_house <- list()
# 
# for(i in 1:nrow(house_passed)) {
#   new_table <- crawl_function(house_passed$cosponsor_url[i])
#   cosponsor_list_house[[length(cosponsor_list_house) + 1]] <- new_table
# }
# 
# signed <- rbindlist(cosponsor_list_signed,use.names = FALSE)
# senate <- rbindlist(cosponsor_list_senate,use.names = FALSE)
# house <- rbindlist(cosponsor_list_house,use.names = FALSE)
# 
# cosponsors <- rbind(signed, senate, use.names=FALSE)
# cosponsors <- rbind(cosponsors, house, use.names = FALSE)

### save as data.frame output so only have to do this once ###
# write.csv(cosponsors, "~/Desktop/my_projects/personal_site/content/moment/cosponsors.csv")
```

```{r manipulate data, echo = FALSE, warning = FALSE, message = FALSE}
# manipulate cosponsors
cosponsors <- read_csv("~/Desktop/my_projects/personal_site/content/moment/cosponsors.csv") %>%
                         select(-1)
cosponsors <- cosponsors %>% 
  mutate(bill = sub(" -.*","", cosponsors$bill)) %>%
  rename(cosponsor = Cosponsor)

cosponsors$bill <- gsub("\\.", "", cosponsors$bill)
cosponsors$cosponsor <- gsub("\\*", "", cosponsors$cosponsor)

# clean bills for join
bills$legislation_number <- gsub("\\.", "", bills$legislation_number)
bills$legislation_number <- gsub(" ", "", bills$legislation_number)

# join for summary table
summary <- left_join(bills, cosponsors, by = c("legislation_number" = "bill")) %>% distinct() %>%
  select(-2,-3,-9)

# fix doug jones duplicate
for(i in 1:nrow(summary)){
  if(summary$cosponsor[i] == "Sen. Jones, Doug  [D-AL]"){
    summary$cosponsor[i] <- "Sen. Jones, Doug [D-AL]"
  }
  if(summary$sponsor[i]== "Sen. Jones, Doug  [D-AL]"){
    summary$sponsor[i] <- "Sen. Jones, Doug [D-AL]"
  }
}
```
  
The 116th Congress of the United States convened on January 3rd, 2019 and ended on January 3rd, 2021, giving way to the 117th Congress. Before looking forward to the accomplishments of the next Congress, it's worth looking back at the record of the 116th Congress's legislative accomplishments and the elected officials that made Washington work. 

Over 15,000 bills were introduced in the House and Senate, 1,800 bills passed at least one chamber of Congress, 344 bills became law, and President Trump vetoed 10 bills.

```{r sankey, echo = FALSE, message=FALSE, warning=FALSE,fig.align='center', fig.height= 8}
# make sankey of overall bills - too messy for legislative
bill_flow_edges <- read_csv("bill_flow_116_edges.csv") 
bill_flow_nodes <- read_csv("bill_flow_116_nodes.csv")

bill_flow_edges <- bill_flow_edges %>%
  left_join(bill_flow_nodes, by = c("source" = "label")) %>% 
  rename(from = id)

bill_flow_edges <- bill_flow_edges %>% 
  left_join(bill_flow_nodes, by = c("destination" = "label")) %>% 
  rename(to = id)

bill_flow_edges <- select(bill_flow_edges, from, to, weight)

# manipulate nodes and edges for d3 specification
bill_flow_nodes_d3 <- mutate(bill_flow_nodes, id = id - 1)
bill_flow_edges_d3 <- mutate(bill_flow_edges, from = from - 1, to = to - 1)

# prepare colour scale
sankey_color <-'d3.scaleOrdinal() .range(["#3395FF","#EBE965","#AAEB65","green","red","#3395FF","orange"])'

sankeyNetwork(Links = bill_flow_edges_d3, Nodes = bill_flow_nodes_d3, Source = "from", Target = "to",
              NodeID = "label", Value = "weight", fontSize = 16,
              sinksRight = F,
              iterations = 1000,
              nodeWidth = 30,
              nodePadding = 100,
              colourScale = sankey_color)
```

***  

Before moving onto a full network analysis of the 116th Congress, please find below some top-line metrics from the 116th Congress situated in comparative historical context.

```{r historical, echo = FALSE, message=FALSE, warning=FALSE,fig.align='center', fig.height= 8}
# historical bar chart of number of bills passed per congress
historical_bar <- read_csv("historical_congress_bills.csv") %>%
  mutate(Congress = factor(Congress, levels = Congress))

historical_bar_longer <- historical_bar %>%
  pivot_longer(cols = 2:4, names_to = "metric", values_to = "values")

ggplot(historical_bar_longer, aes(x = Congress, y = values, fill = `Control of Government`)) +
  
  geom_col(alpha = .9, 
           size = 1) + 
  
  facet_wrap(~metric, 
             ncol = 1, 
             scales = "free") +

  theme_tufte() +
  
  scale_x_discrete(limits = rev(levels(historical_bar_longer$Congress))) +
  
  scale_y_continuous(n.breaks = 3) +
  
  labs(x = "Session of Congress",
       y = NULL,
       title = "Historical legislative activity by Congress and the President") +

  theme(plot.caption = element_text(size = 12, face = "italic"),
        plot.title = element_text(size=14, face = "bold", hjust = .5 ),
        plot.subtitle = element_text(size=12),
        text=element_text(family="Georgia"),
        axis.title.y = element_text(size = 12),
        axis.text = element_text(size = 14),
        axis.text.x = element_text(angle = 25, vjust = 1, hjust = 1, size = 8),
        panel.grid.major.y = element_line(colour = "grey", size = .5),
        panel.grid.minor.y = element_line(colour = "grey90", size = 0.3),
        legend.position = "top",
        strip.text.x = element_text(size = 14, colour = "black")) 

```
 
The number of bills considered on the floor of both chambers of Congress has remained somewhat steady, but **the number of bills passed into law has decreased by more than half over the past 40 years.** The presidential veto power has been used more sparingly in recent decades, and almost exclusively during times of divided government - President Carter (1977-80) is a clear outlier in this respect.  

***  
## Senate Legislative Network Analysis

**Now, for the main subject of this article: the legislative network of Congress.** The Senate and House are groups of individual senators and representatives that, taken together, form the Congress of United States. Much like a social network like Instagram or the human nervous system, representatives in Congress are connected through various relationships, such as political party or home state.

The primary role of senators and representatives is to legislate: to write and pass bills that ultimately become law. For those that need it, [click here for a quick refresher](https://www.youtube.com/watch?v=FFroMQlKiag) on the legislative process. But senators and representatives don't write and pass bills alone - they need to work with other elected officials to shepherd bills through committee, garner support for a floor vote, and pass their respective chambers. 

These legislative relationships are best represented through the [**sponsor-cosponsor relationship**](https://www.legion.org/legislative/thomas/8974/part-4-sponsors-and-co-sponsors):    
1. For a bill to be submitted for consideration, at least one senator or representative must sponsor it. The first name on the bill is the sponsor, and all additional names are known as "original" cosponsors.  
2. The sponsor of a bill can then recruit others to cosponsor the bill after introduction, indicating support and generating momentum.  
3. A bill can have any number of cosponsors, and generally many more representatives will ultimately vote for a bill than cosponsor it.   

Network analysis allows for a quantitative analysis of this legislative network. In a network, **nodes** are connected to each other through **ties** of varying directions and strengths. In this legislative network, ***the senators and representatives are nodes connected to each other through the bills they have sponsored and cosponsored***. 


```{r senate graph, echo = FALSE, message=FALSE, warning=FALSE,fig.align='center', fig.height= 8}
###### SENATE NETWORK #######
# prepare summary table for network analysis (https://www.jessesadler.com/post/network-analysis-with-r/)
# let's start with just bills that became law from senate
passed_summary <- summary %>%
  filter(status == "Signed into law") %>%
  filter(str_detect(legislation_number,"S"))
# first create node list
sponsors <- passed_summary %>%
  distinct(sponsor) %>%
  rename(label = sponsor)

cosponsors <- passed_summary %>%
  distinct(cosponsor) %>%
  rename(label = cosponsor) %>%
  filter(label!= "no sponsor")

# join together to assign unique ids and create nodes df
nodes <- full_join(sponsors, cosponsors, by = "label")
nodes <- nodes %>% 
  rowid_to_column("id")

# add partisanship to nodes
for(i in 1:nrow(nodes)){
  if(str_detect(nodes$label[i], "R-")){
    nodes$group[i] <- "Republican"
  } else if (str_detect(nodes$label[i], "D-")){
    nodes$group[i] <- "Democrat"
  } 
  else nodes$group[i] <- "Independent"
}

# add color to nodes
for(i in 1:nrow(nodes)){
  if(nodes$group[i] == "Republican"){
    nodes$color[i] <- "red"
  } else if (nodes$group[i] == "Democrat"){
    nodes$color[i] <- "blue"
  } 
  else nodes$color[i] <- "green"
}

# add title to nodes
for(i in 1:nrow(nodes)){
  nodes$title[i] <- nodes$label[i]
}
# nodes <- nodes %>%
#   mutate(label = gsub(",.*$", "", label),
#          title = gsub(",.*$", "", title)) %>%
#   mutate(label = sub('.*\\.', "", label),
#          title = sub('.*\\.', "", title)) %>%
#   mutate(label = trimws(label),
#          title = trimws(title))

# create edge list
edges <- passed_summary %>%
  filter(cosponsor!= "no sponsor") %>%
  # mutate(sponsor = gsub(",.*$", "", sponsor),
  #        cosponsor = gsub(",.*$", "", cosponsor)) %>%
  # mutate(sponsor = sub('.*\\.', "", sponsor),
  #        cosponsor = sub('.*\\.', "", cosponsor)) %>%
  # mutate(sponsor = trimws(sponsor),
  #        cosponsor = trimws(cosponsor))%>%
  group_by(sponsor, cosponsor) %>%
  summarise(weight = n()) %>%
  ungroup()

# join together to finish edges df
edges <- edges %>%
  left_join(nodes, by = c("sponsor" = "label")) %>%
  rename(from = id)

edges <- edges %>%
  left_join(nodes, by = c("cosponsor" = "label")) %>%
  rename(to = id)

edges <- select(edges, from, to, weight)

# now nodes and edges df are prepped
bills_passed_igraph <- graph_from_data_frame(d = edges, vertices = nodes, directed = TRUE)

# compute weighted betweenness 
dist_weight = 1/E(bills_passed_igraph)$weight
E(bills_passed_igraph)$betweenness <- edge_betweenness(bills_passed_igraph, weights = dist_weight)

# and median betweenness
median_betweenness = median(E(bills_passed_igraph)$betweenness)

# # static graph
# ggraph(bills_passed_igraph, layout = "with_kk") +
#   geom_edge_link(aes(alpha = betweenness, filter = betweenness > median_betweenness)) +
#   geom_node_point(aes(color = factor(party))) +
#   geom_node_text(aes(label = label), repel = TRUE) 
  

# draft interactive
vis_data <- toVisNetworkData(bills_passed_igraph)

visNetwork(
  nodes = vis_data$nodes, 
  edges = vis_data$edges,
  main = "Senate Legislative Network: 116th Congress"
) %>%
# Can set specific layout with visIgraphLayout()
visIgraphLayout(layout = "layout_with_kk") %>%
visGroups(groupname = "Democrat", shape = "icon",
          icon = list(code = "f007", color = "blue", size = 100))%>% 
  
visGroups(groupname = "Republican", shape = "icon",
            icon = list(code = "f007", color = "red", size = 100)) %>% 
  
visGroups(groupname = "Independent", shape = "icon",
            icon = list(code = "f007", color = "green", size = 100))%>% 
  
visEdges(color = list(color = "grey", highlight = "purple"),
         smooth = FALSE,
         dashes = FALSE,
         shadow = FALSE,
         arrows = "from") %>%
  addFontAwesome() %>% # required for icons

visOptions(highlightNearest = list(enabled = T, hover = T),
           selectedBy = "title") %>%
  
visInteraction(zoomView = FALSE,
               navigationButtons = TRUE) %>%
  
visLayout(randomSeed = 123)

```
  
  
  
The Senate legislative network above visualizes the sponsor-cosponsor relationship between senators for every bill that became law in the 116th Congress. The direction of the arrows illustrate the flow of cosponsorships from cosponsors to the original sponsor of a bill. **Senators at the center of the network worked together more closely with a higher number of other senators than those at the further edges of the network.** Clicking a particular senator's node will highlight every senator they worked with to pass legislation in the 116th Congress.
  
***  
  
Below, the chord diagram represents the sponsor-cosponsor relationship between Republicans, Democrats and the two Independents in the Senate. 
  
```{r senate chord diagram,echo = FALSE, message=FALSE, warning=FALSE,fig.align='center'}
# senate chord diagram using circulize() package
senate_adj_matrix <- passed_summary %>%
  select(sponsor, cosponsor) %>%
  filter(cosponsor!= "no sponsor") 

# add partisanship to sponsors
for(i in 1:nrow(senate_adj_matrix)){
  if(str_detect(senate_adj_matrix$sponsor[i], "R-")){
    senate_adj_matrix$sponsor_party[i] <- "GOP Sponsor"
  } else if (str_detect(senate_adj_matrix$sponsor[i], "D-")){
    senate_adj_matrix$sponsor_party[i] <- "Dem Sponsor"
  } 
  else senate_adj_matrix$sponsor_party[i] <- "Ind. Sponsor"
}

# add partisanship to cosponsors
for(i in 1:nrow(senate_adj_matrix)){
  if(str_detect(senate_adj_matrix$cosponsor[i], "R-")){
    senate_adj_matrix$cosponsor_party[i] <- "GOP Cosponsor"
  } else if (str_detect(senate_adj_matrix$cosponsor[i], "D-")){
    senate_adj_matrix$cosponsor_party[i] <- "Dem Cosponsor"
  } 
  else senate_adj_matrix$cosponsor_party[i] <- "Ind. Cosponsor"
}

# create adjacency matrix
#senate_adj_matrix <- as_adjacency_matrix(bills_passed_igraph, attr="weight", sparse = FALSE)
senate_adj_matrix <- senate_adj_matrix %>%
  select(sponsor_party,cosponsor_party) %>%
  arrange(cosponsor_party) %>%
  graph_from_data_frame() %>%
  as_adjacency_matrix(sparse = FALSE)

circos.clear()
# set gap
circos.par(gap.after = 20, start.degree = -125)
# set colors
grid.col = c( "GOP Sponsor" = "red",  "Dem Sponsor" = "blue",  "Ind. Sponsor"= "green",
              "GOP Cosponsor" = "red",  "Dem Cosponsor" = "blue",  "Ind. Cosponsor"= "green")

link.col = c( "purple", "red", "green", "blue", "purple", "green")

chordDiagram(t(senate_adj_matrix),
             grid.col = grid.col,
             col = link.col,
             big.gap = 20,
             annotationTrack = c("name","grid"),
             annotationTrackHeight = c(.05,.05),
             order = sort(rownames(senate_adj_matrix)),
             directional = 1,
             direction.type = c("diffHeight", "arrows"),
             link.arr.type = "triangle",
             diffHeight  = -0.05)

```
  
Similar to the full network visualization prior, the arrows indicate the flow of cosponsorships from cosponsors to original bill sponsors. Though a healthy majority of bills passed were originally sponsored by a Republican (the GOP had Senate control in the 116th Congress), a fair number of passed legislation was sponsored by Democrats. Given the current (60 vote threshold in the Senate to overcome the filibuster)[https://www.brookings.edu/policy2020/votervital/what-is-the-senate-filibuster-and-what-would-it-take-to-eliminate-it/] and pass legislation, it's not surprising a significant number of Democrats cosponsored Republican bills that became law, and vice-versa. This will look different in the majority-rule house further below.  
  
***  
## The Influence of Individual Senators   

**Is it possible to know which senators were most important to the legislative process in the 116th Congress?** It depends on your definition of "important" - most bills passed? The price tag of bills? If a bill was related to certain critical policy areas, such as national security or defense? In network analysis, the concept of centrality attempts to determine the importance of nodes. *As there are various definitions of importance, there are various measurements of centrality.*

**Some standard centrality measurements include:**  
-  **degree**: how many direct ties a specific node has to other nodes  
-  **strength**: the strength of the ties between a specific node and other nodes  
-  **betweenness**: how often a specific node lies on the shortest path between other nodes, i.e. the influence of a specific node across the entire network  

**In the case of this legislative network:**   
-  **degree**: how many cosponsorships a specific senator recruited or extended  
-  **strength**: the strength of the ties between a specific senator and other senators   
-  **betweenness**: the influence of a specific senator across the entire legislative network  

```{r senate stats table, echo = FALSE, message=FALSE, warning=FALSE,fig.align='center'}
# senate centrality
senate_nodes <- nodes %>% select(1,2)
senate_edges <- edges %>% select(1:3)

senate_g <- graph_from_data_frame(senate_edges, directed = TRUE, vertices = senate_nodes)

senate_nodes <- senate_nodes %>%
  mutate(degree = igraph::degree(senate_g),
         strength = igraph::strength(senate_g),
         betweenness = igraph::betweenness(senate_g),
         closeness = igraph::closeness(senate_g))

# senate bills sponsored and cosponsored
senator_stats_sponsor <- passed_summary %>%
  select(sponsor,title) %>%
  distinct() %>%
  group_by(sponsor) %>%
  summarize(bills_sponsored = n())

senator_stats_cosponsor <- passed_summary %>%
  select(cosponsor,title) %>%
  distinct() %>%
  group_by(cosponsor) %>%
  summarize(bills_cosponsored = n())

senator_stats <- full_join(senator_stats_sponsor,senator_stats_cosponsor, by = c("sponsor" = "cosponsor")) %>%
  filter(sponsor != "no sponsor") 

# make senate table
senate_table <- senate_nodes %>%
  select(2:6) %>%
  rename(Senator = label) %>%
  mutate(betweenness = round(betweenness,0),
         closeness = round(closeness,4)) %>%
  left_join(senator_stats, by = c("Senator" = "sponsor")) %>%
  rename(`passed bills sponsored` = "bills_sponsored",
         `passed bills cosponsored` = "bills_cosponsored") %>%
  select(1,6,7,2:4) %>%
  replace_na(list("passed bills sponsored" = 0,
                  "passed bills cosponsored" = 0))

reactable(senate_table,
          defaultSorted = list("betweenness" = "desc"),
          searchable = TRUE,
          bordered = TRUE,
          striped = TRUE,
          highlight = TRUE,
          compact = TRUE, 
          resizable = TRUE,
          columns = list(Senator = colDef(width = 250),
                         betweenness = colDef(width = 140),
                         degree = colDef(width = 100),
                         strength = colDef(width = 100)))

```
  
***
## House Legislative Network Analysis  

The same set of analyses as above, but for the House of Representatives.  

```{r house graph, echo = FALSE, message=FALSE, warning=FALSE,fig.align='center', fig.height= 10}
###### House NETWORK #######
# prepare summary table for network analysis (https://www.jessesadler.com/post/network-analysis-with-r/)
# let's start with just bills that became law from senate
house_summary <- summary %>%
  filter(status == "Signed into law") %>%
  filter(str_detect(legislation_number,"H"))
# first create node list
house_sponsors <- house_summary %>%
  distinct(sponsor) %>%
  rename(label = sponsor)

house_cosponsors <- house_summary %>%
  distinct(cosponsor) %>%
  rename(label = cosponsor) %>%
  filter(label!= "no sponsor")

# join together to assign unique ids and create nodes df
house_nodes <- full_join(house_sponsors, house_cosponsors, by = "label")
house_nodes <- house_nodes %>% 
  rowid_to_column("id")

# add partisanship to nodes
for(i in 1:nrow(house_nodes)){
  if(str_detect(house_nodes$label[i], "R-")){
    house_nodes$group[i] <- "Republican"
  } else if (str_detect(house_nodes$label[i], "D-")){
    house_nodes$group[i] <- "Democrat"
  } 
  else house_nodes$group[i] <- "Independent"
}

# add color to nodes
for(i in 1:nrow(house_nodes)){
  if(house_nodes$group[i] == "Republican"){
    house_nodes$color[i] <- "red"
  } else if (house_nodes$group[i] == "Democrat"){
    house_nodes$color[i] <- "blue"
  } 
  else house_nodes$color[i] <- "green"
}

# add title to nodes
for(i in 1:nrow(house_nodes)){
  house_nodes$title[i] <- house_nodes$label[i]
}
# create edge list
house_edges <- house_summary %>%
  filter(cosponsor!= "no sponsor") %>%
  group_by(sponsor, cosponsor) %>%
  summarise(weight = n()) %>%
  ungroup()

# join together to finish edges df
house_edges <- house_edges %>%
  left_join(house_nodes, by = c("sponsor" = "label")) %>%
  rename(from = id)

house_edges <- house_edges %>%
  left_join(house_nodes, by = c("cosponsor" = "label")) %>%
  rename(to = id)

house_edges <- select(house_edges, from, to, weight)

# now nodes and edges df are prepped
house_bills_passed_igraph <- graph_from_data_frame(d = house_edges, vertices = house_nodes, directed = TRUE)

# compute weighted betweenness 
house_dist_weight = 1/E(house_bills_passed_igraph)$weight
E(house_bills_passed_igraph)$betweenness <- edge_betweenness(house_bills_passed_igraph, weights = house_dist_weight)

# and median betweenness
house_median_betweenness = median(E(house_bills_passed_igraph)$betweenness)

# # static graph
# ggraph(bills_passed_igraph, layout = "with_kk") +
#   geom_edge_link(aes(alpha = betweenness, filter = betweenness > median_betweenness)) +
#   geom_node_point(aes(color = factor(party))) +
#   geom_node_text(aes(label = label), repel = TRUE) 


# draft interactive
house_vis_data <- toVisNetworkData(house_bills_passed_igraph)

visNetwork(
  nodes = house_vis_data$nodes, 
  edges = house_vis_data$edges,
  main = "House Legislative Network: 116th Congress"
) %>%
  # Can set specific layout with visIgraphLayout()
  visIgraphLayout(layout = "layout_with_kk") %>%
  visGroups(groupname = "Democrat", shape = "icon",
            icon = list(code = "f007", color = "blue", size = 100))%>% 
  
  visGroups(groupname = "Republican", shape = "icon",
            icon = list(code = "f007", color = "red", size = 100)) %>% 
  
  visGroups(groupname = "Independent", shape = "icon",
            icon = list(code = "f007", color = "green", size = 100))%>% 
  
  visEdges(color = list(color = "grey", highlight = "purple"),
           smooth = FALSE,
           dashes = FALSE,
           shadow = FALSE,
           arrows = "from") %>%
  addFontAwesome() %>% # required for icons
  
  visOptions(highlightNearest = list(enabled = T, hover = T),
             selectedBy = "title") %>%
  
  visInteraction(zoomView = FALSE,
                 navigationButtons = TRUE) %>%
  
  visLayout(randomSeed = 123)
```
  
  
  
***  
  
**In the majority-rule House**, the number of bills that passed were sponsored by Democrats significantly more so than by Republicans (the Democrats regained control for the 116th Congress).  

```{r house chord diagram,echo = FALSE, message=FALSE, warning=FALSE,fig.align='center'}
# house chord diagram using circulize() package
house_adj_matrix <- house_summary %>%
  select(sponsor, cosponsor) %>%
  filter(cosponsor!= "no sponsor") 

# add partisanship to sponsors
for(i in 1:nrow(house_adj_matrix)){
  if(str_detect(house_adj_matrix$sponsor[i], "R-")){
    house_adj_matrix$sponsor_party[i] <- "GOP Sponsor"
  } else if (str_detect(house_adj_matrix$sponsor[i], "D-")){
    house_adj_matrix$sponsor_party[i] <- "Dem Sponsor"
  } 
  else house_adj_matrix$sponsor_party[i] <- "Ind. Sponsor"
}

# add partisanship to cosponsors
for(i in 1:nrow(house_adj_matrix)){
  if(str_detect(house_adj_matrix$cosponsor[i], "R-")){
    house_adj_matrix$cosponsor_party[i] <- "GOP Cosponsor"
  } else if (str_detect(house_adj_matrix$cosponsor[i], "D-")){
    house_adj_matrix$cosponsor_party[i] <- "Dem Cosponsor"
  } 
  else house_adj_matrix$cosponsor_party[i] <- "Ind. Cosponsor"
}

# create adjacency matrix
house_adj_matrix <- house_adj_matrix %>%
  select(sponsor_party,cosponsor_party) %>%
  arrange(cosponsor_party) %>%
  graph_from_data_frame() %>%
  as_adjacency_matrix(sparse = FALSE)

circos.clear()
# set gap
circos.par(gap.after = 30, start.degree = 92)
# set colors
grid.col = c( "GOP Sponsor" = "red",  "Dem Sponsor" = "blue",  "Ind. Sponsor"= "green",
              "GOP Cosponsor" = "red",  "Dem Cosponsor" = "blue",  "Ind. Cosponsor"= "green")

link.col = c( "blue", "purple", "green", "purple", "red", "green")

chordDiagram(t(house_adj_matrix),
             grid.col = grid.col,
             col = link.col,
             annotationTrack = c("name","grid"),
             annotationTrackHeight = c(.05,.05),
             order = sort(rownames(senate_adj_matrix)),
             directional = 1,
             direction.type = c("diffHeight", "arrows"),
             link.arr.type = "triangle",
             diffHeight  = -0.05)




```
  
## House Summary Stats and Centality Measurements  
  
Can you find your representative?  

```{r house stats table, echo = FALSE, message=FALSE, warning=FALSE,fig.align='center'}

# house centrality
house_nodes <- house_nodes %>% select(1,2)
house_edges <- house_edges %>% select(1:3)

house_g <- graph_from_data_frame(house_edges, directed = TRUE, vertices = house_nodes)

house_nodes <- house_nodes %>%
  mutate(degree = igraph::degree(house_g),
         strength = igraph::strength(house_g),
         betweenness = igraph::betweenness(house_g),
         closeness = igraph::closeness(house_g))

# house bills sponsored and cosponsored
house_stats_sponsor <- house_summary %>%
  select(sponsor,title) %>%
  distinct() %>%
  group_by(sponsor) %>%
  summarize(bills_sponsored = n())

house_stats_cosponsor <- house_summary %>%
  select(cosponsor,title) %>%
  distinct() %>%
  group_by(cosponsor) %>%
  summarize(bills_cosponsored = n())

house_stats <- full_join(house_stats_sponsor,house_stats_cosponsor, by = c("sponsor" = "cosponsor")) %>%
  filter(sponsor != "no sponsor") 

# make house table
house_table <- house_nodes %>%
  select(2:6) %>%
  rename(Congressperson = label) %>%
  mutate(betweenness = round(betweenness,0),
         closeness = round(closeness,4)) %>%
  left_join(house_stats, by = c("Congressperson" = "sponsor")) %>%
  rename(`passed bills sponsored` = "bills_sponsored",
         `passed bills cosponsored` = "bills_cosponsored") %>%
  select(1,6,7,2:4) %>%
  replace_na(list("passed bills sponsored" = 0,
                  "passed bills cosponsored" = 0))


reactable(house_table,
          defaultSorted = list("betweenness" = "desc"),
          searchable = TRUE,
          bordered = TRUE,
          striped = TRUE,
          highlight = TRUE,
          compact = TRUE, 
          resizable = TRUE,
          columns = list(Congressperson = colDef(width = 250),
                          betweenness = colDef(width = 140),
                         degree = colDef(width = 100),
                         strength = colDef(width = 100)))


```


```{r similarity senate, echo = FALSE, message=FALSE, warning=FALSE,fig.align='center'}
# full_senate <- summary %>%
#   filter(str_detect(legislation_number,"S"))
# 
# # first create node list
# full_senate_sponsors <- full_senate %>%
#   distinct(sponsor) %>%
#   rename(label = sponsor)
# 
# full_senate_cosponsors <- full_senate %>%
#   distinct(cosponsor) %>%
#   rename(label = cosponsor) %>%
#   filter(label!= "no sponsor")
# 
# # join together to assign unique ids and create nodes df
# full_senate_nodes <- full_join(full_senate_sponsors, full_senate_cosponsors, by = "label")
# full_senate_nodes <- full_senate_nodes %>% 
#   rowid_to_column("id")
# 
# # create edge list
# full_senate_edges <- full_senate %>%
#   filter(cosponsor!= "no sponsor") %>%
#   group_by(sponsor, cosponsor) %>%
#   summarise(weight = n()) %>%
#   ungroup()
# 
# # join together to finish edges df
# full_senate_edges <- full_senate_edges %>%
#   left_join(full_senate_nodes, by = c("sponsor" = "label")) %>%
#   rename(from = id)
# 
# full_senate_edges <- full_senate_edges %>%
#   left_join(full_senate_nodes, by = c("cosponsor" = "label")) %>%
#   rename(to = id)
# 
# full_senate_edges <- select(edges, from, to, weight)
# 
# # create full senate graph
# full_senate_g <- graph_from_data_frame(full_senate_edges, directed = TRUE, vertices = full_senate_nodes)
# 
# # add full senate centrality
# full_senate_nodes <- full_senate_nodes %>%
#   mutate(degree = igraph::degree(full_senate_g),
#          strength = igraph::strength(full_senate_g),
#          betweenness = igraph::betweenness(full_senate_g),
#          closeness = igraph::closeness(full_senate_g))
# 
# # create senate adjacency matrix
# senate_adj <- as_adjacency_matrix(full_senate_g, attr = "weight", names = TRUE, sparse = FALSE)
# 
# # create senate similarity matrix
# cor_senate <- cor(senate_adj)
# cor_senate <- cor_senate[,c(-102,-103)]
# cor_senate <- cor_senate[c(-102,-103),]
# diag(cor_senate) <- 0
# 
# # visualize similarity
# # Flatten S to be a vector
# flat_cor_senate <- as.vector(cor_senate)
# 
# # Plot a histogram of similarities
# hist(flat_cor_senate, xlab = "Similarity", main = "Histogram of Senate similarity")
# 
# # visualize other correlations of senate centrality measurements
# # degree vs strength
# ggplot(senate_nodes, aes(x = degree, y = strength)) +
#   # Add a point geom
#   geom_point() +
#   # Add a smooth geom with linear regression method
#   geom_smooth(method = "lm", se = FALSE)
# 
# # Calculate correlation 
# cor_sen_degree_strength <- cor(full_senate_nodes$degree, full_senate_nodes$strength)
# 
# # degree vs betweenness
# ggplot(full_senate_nodes, aes(x = degree, y = betweenness)) +
#   # Add a point geom
#   geom_point() +
#   # Add a smooth geom with linear regression method
#   geom_smooth(method = "lm", se = FALSE)
# 
# # Calculate correlation 
# cor_sen_degree_betweenness <- cor(full_senate_nodes$degree, full_senate_nodes$betweenness)
# 
# # strength vs betweenness
# ggplot(full_senate_nodes, aes(x = strength, y = betweenness)) +
#   # Add a point geom
#   geom_point() +
#   # Add a smooth geom with linear regression method
#   geom_smooth(method = "lm", se = FALSE)
# 
# # Calculate correlation 
# cor_sen_strength_betweenness <- cor(full_senate_nodes$strength, full_senate_nodes$betweenness)
# 
# # # Convert weighted similarity matrix to a graph
# # sim_senate <- graph_from_adjacency_matrix(cor_senate, mode = "directed", weighted = TRUE)
# # 
# # # Convert sim_senate to a data.frame
# # sim_senate_df <- igraph::as_data_frame(sim_senate, what = "edges")
# # 
# # # Convert sim_senate_df to a tibble
# # sim_senate_tib <- as_tibble(sim_senate_df)
# # 
# # # join to senate nodes
# # senate_nodes <- senate_nodes %>% mutate(id = as.character(id))
# # sim_senate_joined <- sim_senate_tib %>% 
# #   # Left join to nodes matching "from" to "id"
# #   left_join(senate_nodes, by = c("from" = "id")) %>% 
# #   # Left join to nodes matching "to" to "id", setting suffixes
# #   left_join(senate_nodes, by = c("to" = "id"), suffix = c("_from", "_to")) %>%
# #   rename(similarity = weight)
# # 
# # sim_senate_joined %>% 
# #   filter(degree_from >= median(degree_from),
# #          degree_to >= median(degree_to)) %>%
# # arrange(desc(similarity))
# # 
# # sim_senate_filtered <- sim_senate_joined %>% 
# #   # Filter on similarity greater than 0.6
# #   filter(similarity > 0.8)
# # 
# # # Convert to an directed graph
# # sim_senate_filtered_network <- graph_from_data_frame(sim_senate_filtered, directed = TRUE)
# # 
# # V(sim_senate_filtered_network)$name <- sim_senate_filtered$label_to
# # # Plot with Kamada-Kawai layout
# # ggraph(sim_senate_filtered_network, layout = "with_kk") + 
# #   # Add an edge link geom, mapping transparency to similarity
# #   geom_edge_link(aes(alpha = similarity)) +
# #   geom_node_text(aes(label = name))
```

```{r clustered senate similarity network, echo = FALSE, message=FALSE, warning=FALSE,fig.align='center'}
# distance_sen_network <- 1 - cor_senate
# 
# d <- as.dist(distance_sen_network)
# 
# # run average-linkage clustering method and plot the dendrogram 
# senate_dendogram <- stats::hclust(d, method = "average")
# plot(senate_dendogram)
# 
# # Cut the dendrogram tree into 4 clusters
# cls <- cutree(senate_dendogram, k = 4)
# 
# # add to senate nodes
# full_senate_nodes_clustered <- full_senate_nodes %>%
#   filter(id <= 101) %>%
#   mutate(cluster = cls)
# 
# # create clustered graph
# clustered_senate_graph <- graph_from_data_frame(full_senate_edges, directed = TRUE, vertices = full_senate_nodes_clustered)
# 
# # Update the plot
# ggraph(clustered_senate_graph, layout = "with_kk") + 
#   geom_edge_link(aes(alpha = weight), show.legend=FALSE) +  
#   geom_node_point(aes(color = factor(cluster))) + 
#   geom_node_text(aes(label = label)) +
#   labs(color = "cluster") + 
#   # Facet the nodes by cluster, with a free scale
#   facet_nodes(~ cluster, scales="free") 

```

## Reference - All Bills Passed in 116th Congress  

```{r all bills, echo = FALSE, message=FALSE, warning=FALSE,fig.align='center'}
all_bills <- bills %>%
  filter(status == "Signed into law") %>%
  select(1,4,5,6,7,2) %>%
  distinct() %>%
  arrange(title) %>%
  rename(`Legislation #` = legislation_number,
         Title = title,
         Sponsor = sponsor,
         `Cosponsors` = number_of_cosponsors,
         Committees = committees,
         `Bill URL` = url )

reactable(all_bills,
          searchable = TRUE,
          bordered = TRUE,
          striped = TRUE,
          highlight = TRUE,
          compact = TRUE, 
          resizable = TRUE,
          columns = list("Legislation #" = colDef(width = 100),
                         Title = colDef(width = 300)))
  #                                       cell = function() {
  #   # Render as a link
  #   url <- all_bills$`Bill URL`
  #   htmltools::tags$a(href = url, target = all_bills$title)
  # })))
```
  
***  

**Questions or comments?** *Please reach out!*  
  
**Data Source:** [congress.gov](https://www.congress.gov/)  
  
*Created in R.*



