---
title: "Which Democratic candidates are getting mentioned the most on cable and online news?"
author: "Tanner Gildea"
date: "2019-07-18"
output: html_document
tags:
  - '2020'
---


FiveThirtyEight is currently running an [interesting series](https://fivethirtyeight.com/tag/media/) analyzing how cable news and online media sources are covering the Democratic presidential candidates. [Dhrumil Mehta](https://fivethirtyeight.com/contributors/dhrumil-mehta/), one of FiveThirtyEight's data journalists, has  diligently kept track of how coverage of individual candidates has risen and fallen week to week. They have also kindly shared the underlying data [here](https://github.com/fivethirtyeight/data/tree/master/media-mentions-2020).

Personally, I was interested in seeing how coverage of the top candidates has changed since the beginning of the race (roughly January of 2019). I also wanted to see if coverage of candidates diverged based on media type - cable or online. 

The graph below helps answer these two questions.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(janitor)
library(readxl)
library(gt)
library(ggthemes)
library(data.table)


download.file(url = "https://raw.githubusercontent.com/fivethirtyeight/data/master/media-mentions-2020/cable_weekly.csv",
              destfile  = "~/Desktop/R.projects/personal_site/content/moment/cable_weekly.csv",
              mode = "wb")    

download.file(url = "https://raw.githubusercontent.com/fivethirtyeight/data/master/media-mentions-2020/online_weekly.csv",
              destfile  = "~/Desktop/R.projects/personal_site/content/moment/online_weekly.csv",
              mode = "wb")   

cable <- read_csv("cable_weekly.csv")

online <- read_csv("online_weekly.csv")

fs::file_delete(c("cable_weekly.csv", "online_weekly.csv"))



cable_joined_online <- merge(cable, online, by = c("name","date" ))
```

```{r join, fig.width = 10, fig.height = 8}
clean_cable_joined_online <- cable_joined_online %>% 
  filter(name %in% c("Joe Biden", 
                     "Pete Buttigieg", 
                     "Elizabeth Warren", 
                     "Kamala Harris", 
                     "Bernie Sanders",
                     "Beto O'Rourke",
                     "Julian Castro",
                     "Amy Klobuchar",
                     "Cory Booker")) %>% 
  group_by(name)

# Needed to distinguish on graph w/ fill and color

clean_cable_joined_online$query.x <- "Cable News"
clean_cable_joined_online$query.y <- "Online Articles"

# Start graph

ggplot(clean_cable_joined_online, aes(x = date, 
                        y = pct_of_all_candidate_stories,
                        color = query.x,
                        fill = query.x)) +
  
  geom_point() +
  
  geom_smooth(level  = .50, method = "loess") +
  
  # for second line
  
  geom_point(aes(x = date, 
                        y = pct_of_all_candidate_clips,
                        color = query.y,
                        fill = query.y)) +
  
  geom_smooth(aes(x = date, 
                        y = pct_of_all_candidate_clips,
                        color = query.y,
                        fill = query.y),
              
              level  = .50, method = "loess") +
  
  # Makes scale more readable
  
  scale_y_sqrt(labels = scales::percent,
    breaks = c(.01, .025, .05, 0.1, .3, 0.2, 0.4, .5, .6)) +
  
  facet_wrap(~name) +
  
#  scale_x_date(date_breaks = "1 month") + 
  
  # Style Settings
  
  labs(x = NULL, y = "Percent of Articles and Stories That Mention Each Candidate",
       title = "How often are the candidates mentioned in cable and online news?",
       subtitle = "Cable includes CNN, Fox News, and MSNBC - Online articles sourced from the Media Cloud database",
       caption = "Source: FiveThirtyEight.com. Percentages can add to greater than 100% since articles or stories often mention multiple candidates.") +
  
  scale_color_brewer(palette = "Paired") +
  scale_fill_brewer(palette = "Paired") +
  
  scale_fill_discrete(name = "Media Type") + 
  scale_color_discrete(name = "Media Type") +
  
  
  theme_tufte() +
  
  theme(plot.title = element_text(size = rel(1.5), face ="bold"),
        plot.subtitle = element_text(face = "italic"),
        legend.title = element_text(size = rel(1.2), face = "bold"),
        axis.title.y = element_text(size = rel(1.2)),
        plot.caption = element_text(face = "italic"),
        panel.grid.major.y = element_line(color = "gray"),
        axis.line.x = element_line(), axis.line.y = element_line())

```

You can see that for some candidates, e.g. Joe Biden or Beto O'Rourke, cable news and online media coverage tracks relatively closely. Coverage diverges more sharply for other candidates, such as Elizabeth Warren, Bernie Sanders, or Pete Buttigieg.

In general, these nine "higher-tier" candidates are receiving a far higher percentage of the cable news stations' attention than from online sources. Either Joe Biden is gobbling up the internet's attention, or more likely, online articles are slightly more democratized in their coverage of all 24-ish candidates running for the Democratic nomination.

It's important to note that in this race, all news is NOT good news. FiveThirtyEight's dataset only tracks total coverage, not sentiment - good or bad. And increased coverage invites additional scrutiny. 

*****

```{r change_table_cleaning}

# Cable table data

cable_table <- cable %>% 
  select(name, date, pct_of_all_candidate_clips) %>% 
  filter(date == "2019-01-06" | date == "2019-07-07") %>% 
  spread(key = date, value = pct_of_all_candidate_clips) %>% 
  setnames(old=c("2019-01-06","2019-07-07"), new=c("Jan. 6th Cable", "Jul. 7th Cable")) 


cable_table <- cable_table %>% 
  mutate("Percent Cable Change" = cable_table$'Jul. 7th Cable' - cable_table$'Jan. 6th Cable')

# Online table data

online_table <- online %>% 
  select(name, date, pct_of_all_candidate_stories) %>% 
  filter(date == "2019-01-06" | date == "2019-07-07") %>% 
  spread(key = date, value = pct_of_all_candidate_stories) %>% 
  setnames(old=c("2019-01-06","2019-07-07"), new=c("Jan. 6th Online", "Jul. 7th Online"))

online_table <- online_table %>% 
  mutate("Percent Online Change" = online_table$'Jul. 7th Online' - online_table$'Jan. 6th Online')

# join data

joined_table <- left_join(cable_table, online_table) 
```

## Coverage Changes for All "Major" Candidates

According to FiveThirtyEight, there are currently 24 (er, 23, RIP Eric Swalwell) "major candidates" in the Democratic primary field. Although I think there is plenty of room for a distinction to be made between major and minor candidates, FiveThirtyEight uses this categorization to collect data on all of them.

So why not see how their coverage has changed?

```{r change_table, results="asis", fig.width = 5, fig.height = 8}
# construct table
gt(joined_table) %>%
  tab_header(
    title = md("**How much has coverage changed between January and now?**"),
    subtitle = "January 6th to July 7th, 2019") %>% 
  
# Here I simply column names.
  
  cols_label(name = "Candidate",
                            "Jan. 6th Cable" = "Jan. 6",
                            "Jul. 7th Cable" = "Jul. 7th",
                            "Percent Cable Change" = "Percent Change",
                            "Jan. 6th Online" = "Jan. 6th",
                            "Jul. 7th Online" = "Jul. 7th",
                            "Percent Online Change" = "Percent Change") %>%
  
 fmt_percent(columns = vars("Jan. 6th Cable",
                            "Jul. 7th Cable",
                            "Percent Cable Change",
                            "Jan. 6th Online",
                            "Jul. 7th Online",
                            "Percent Online Change"),
             decimals = 1)  %>% 
  
  tab_spanner(label = "Cable Clips", 
              columns = vars("Jan. 6th Cable", "Jul. 7th Cable", "Percent Cable Change")) %>%
  
  tab_spanner(label = "Online Articles", 
              columns = vars("Jan. 6th Online", "Jul. 7th Online", "Percent Online Change")) %>% 
  
    tab_source_note("Source: Internet Archive's Television News Archive via the GDELT Project, Media Cloud, all via FiveThirtyEight.com") %>% 
  
  tab_source_note("Per FiveThirtyEight: “Includes all candidates that qualify as “major” in FiveThirtyEight’s rubric. Each network’s daily news coverage is chopped up into 15-second clips, and each clip that includes a candidate’s name is counted as one mention. For both cable and online news, our search queries look for an exact match for each candidate’s name, except for Julian Castro, for whom our search query is “Julian Castro” OR “Julián Castro”. Media Cloud searches use two of the database’s publication lists: “top online news” and “digital native” publications.”")  %>%  
  
  # add cell coloring
  
  data_color(
    columns = vars("Percent Cable Change"),
    colors = scales::col_quantile(
      palette = c(
        "#FF9999", "#CCCCCC", "#66FF66"),
      domain = NULL)) %>% 

  data_color(
    columns = vars("Percent Online Change"),
    colors = scales::col_quantile(
      palette = c(
        "#FF9999", "#CCCCCC", "#66FF66"),
      domain = NULL)) %>% 
  
  # text aesthetics
  
  tab_options(
    table.width = 750,
    table.font.size = 12,
    column_labels.font.size = 16, 
    column_labels.font.weight = "bold",
    table.border.top.style = "none",
    
    # the magic fix to oversizing!
    
    row.padding = 3)

# This sequence of functions allows the table to transfer nicely into RMarkdown.
  #  as_raw_html() %>% as.character() %>% cat()

```

Candidates who have won the largest increases in coverage include Joe Biden, Kamala Harris, and Pete Buttigieg (who was attracting literally no coverage back in January). Even though he has always attracted significant attention, Biden's front-runner status has led him to dominate coverage since his presidential [campaign announcement](https://time.com/5553949/joe-biden-presidential-campaign-launch-announcement/) in late April.

Other major candidates have trended downwards in their coverage, Beto O'Rourke most precipitously. I would hesistate to read much into Warren's staggering 28 point drop since January - she was the only major candidate announced at the time and was attracting nearly half of the media's attention.

And for the rest of them? The haven't reallymoved much.

```{r debate_bump}

# Cable table data

cable_table_bump <- cable %>% 
  select(name, date, pct_of_all_candidate_clips) %>% 
  filter(date == "2019-06-16" | date == "2019-06-23") %>% 
  spread(key = date, value = pct_of_all_candidate_clips) %>% 
  setnames(old=c("2019-06-16","2019-06-23"), new=c("Before", "After")) 


cable_table_bump <- cable_table_bump %>%
  mutate("Percent Cable Change" = cable_table_bump$'After' - cable_table_bump$'Before')

# Online table data

online_table_bump <- online %>% 
  select(name, date, pct_of_all_candidate_stories) %>% 
  filter(date == "2019-06-16" | date == "2019-06-23") %>% 
  spread(key = date, value = pct_of_all_candidate_stories) %>% 
  setnames(old=c("2019-06-16","2019-06-23"), new=c("Before", "After"))

online_table_bump <- online_table_bump %>%
  mutate("Percent Online Change" = online_table_bump$'After' - online_table_bump$'Before')

# join data

joined_table_bump <- left_join(cable_table_bump, online_table_bump, by = "name") 

```

***** 
## Post-Debate Bump?

The great part about FiveThirtyEight's dataset is that it tracks coverage week to week, so you can isolate changes in coverage around specific events - for example, the first Democratic primary debates. Held over two nights June 26-27 in Miami, FL, the debates saw 20 candidates appear on stage. 

Below is the change in cable and online coverage of candidates based on the news cycle prior to the debate (June 16-22) and the news cycle during and just after (June 23-29).


```{r change_table_bump, results="asis", fig.width = 5, fig.height = 8}
# construct table
gt(joined_table_bump) %>%
  tab_header(
    title = "How did the first debates change candidate coverage?",
    subtitle = "June 16th to June 23rd, 2019") %>% 
  
# Here I simply column names.
  
  cols_label(name = "Candidate",
             "Before.x" = "Before",
             "After.x" = "After",
             "Before.y" = "Before",
             "After.y" = "After",
                            "Percent Cable Change" = "Change",
                            "Percent Online Change" = "Change") %>%
  
 fmt_percent(columns = vars("Before.x",
                            "After.x",
                            "Percent Cable Change",
                            "Before.y",
                            "After.y",
                            "Percent Online Change"), 
             decimals = 1)  %>% 
  
  tab_spanner(label = "Cable Clips", 
              columns = vars("Before.x", "After.x", "Percent Cable Change")) %>%
  
  tab_spanner(label = "Online Articles", 
              columns = vars("Before.y", "After.y", "Percent Online Change")) %>% 
  
  tab_source_note("Source: Same as above.") %>% 
  # add cell coloring
  
  data_color(
    columns = vars("Percent Cable Change"),
    colors = scales::col_quantile(
      palette = c(
        "#FF9999", "#CCCCCC", "#66FF66"),
      domain = NULL)) %>% 

  data_color(
    columns = vars("Percent Online Change"),
    colors = scales::col_quantile(
      palette = c(
        "#FF9999", "#CCCCCC", "#66FF66"),
      domain = NULL)) %>% 
  
    tab_options(
    table.width = 750,
    table.font.size = 12,
    column_labels.font.size = 16, 
    column_labels.font.weight = "bold",
    table.border.top.style = "none",
    
    # the magic fix to oversizing!
    
    row.padding = 3)
  
# This sequence of functions allows the table to transfer nicely into RMarkdown.
    # as_raw_html() %>% as.character() %>% cat()

```


You'll notice that in cable television stories, Kamala Harris clearly had a burst of coverage. A larger pool of candidates received a post-debate bounce online - one that candidates such as Julian Castro badly needed. And some candidates should be worried, especially the single-digit candidates that are holding even or trending down. 

*****
<br />

I realized soon after constructing this table that Mehta had already published [an article](https://fivethirtyeight.com/features/media-outlets-paid-way-more-attention-to-kamala-harris-after-the-debate/) about the effects of the first debate on candidate coverage - makes sense, given this is their full-time job. Maybe I'll beat them to it next time... the [next debates](https://www.cnn.com/politics/live-news/cnn-democratic-debate-draw/index.html) are set for July 30-31 in Detroit!