---
title: "U.S. Covid-19 vaccination live tracker"
author: "Tanner Gildea"
date: "2021-01-28"
output: html_document
tags:
  - "healthcare"
  - "NY Times"
  - "Covid-19"
  - "Tanner Gildea"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(out.width = "120%")
library(tidyverse)
library(rgdal)
library(maptools)
library(mapproj)
library(rgeos)
library(ggplot2)
library(plotly)
library(rjson)
library(ggridges)
library(readr)
library(ggthemes)
library(rvest)
library(httr)
library(sf)
library(psych)
library(lubridate)
library(networkD3)
library(reactable)
library(readbitmap)
library(png)


# download and unzip shape files
download.file("https://www2.census.gov/geo/tiger/GENZ2018/shp/cb_2018_us_state_20m.zip", destfile = "states.zip")
unzip("states.zip") 

# read shapefile
us <- readOGR(dsn = "cb_2018_us_state_20m.shp", layer = "cb_2018_us_state_20m", verbose = FALSE)

# convert it to Albers equal area
us_aea <- spTransform(us, CRS("+proj=laea +lat_0=45 +lon_0=-100 +x_0=0 +y_0=0 +a=6370997 +b=6370997 +units=m +no_defs"))
us_aea@data$id <- rownames(us_aea@data)

# extract, then rotate, shrink & move alaska (and reset projection)
# need to use state IDs via # https://www.census.gov/geo/reference/ansi_statetables.html
alaska <- us_aea[us_aea$STATEFP=="02",]
alaska <- elide(alaska, rotate=-50)
alaska <- elide(alaska, scale=max(apply(bbox(alaska), 1, diff)) / 2.3)
alaska <- elide(alaska, shift=c(-2100000, -2500000))
proj4string(alaska) <- proj4string(us_aea)

# extract, then rotate & shift hawaii
hawaii <- us_aea[us_aea$STATEFP=="15",]
hawaii <- elide(hawaii, rotate=-35)
hawaii <- elide(hawaii, shift=c(5400000, -1400000))
proj4string(hawaii) <- proj4string(us_aea)

# remove old states and put new ones back in; note the different order
# we're also removing puerto rico in this example but you can move it
# between texas and florida via similar methods to the ones we just used
us_aea <- us_aea[!us_aea$STATEFP %in% c("02", "15", "72"),]
us_reshaped <- rbind(us_aea, alaska, hawaii)


#convert shapefile to geoJSON for plotly/leaflet
geojson_file <- "us_reshaped.json"
#writeOGR(us_reshaped, geojson_file,layer = "geojson", driver = "GeoJSON")
#us_json <- fromJSON(file = "us_reshaped.json")

# use glimpse(us_reshaped@data) to identify opportunities to join shapefile with other data
# looks like STUSPS or NAME columns in us_reshaped are good join keys


##################3
###################
###################
# other NYT data

# state population data 2019
download.file("https://www2.census.gov/programs-surveys/popest/tables/2010-2019/state/totals/nst-est2019-01.xlsx", destfile = "state_pop.xlsx")

state_pop <- readxl::read_xlsx("state_pop.xlsx", skip = 2) %>%
  select('Geographic Area',13) %>%
  rename(pop = 2,
         state = 'Geographic Area') %>%
  mutate(state = str_remove_all(state, '\\.'))

file.remove("state_pop.xlsx")

# daily case counts from NYT
daily_states <- read_csv(url("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")) %>%
  group_by(state) %>%
  arrange(state)

daily_states <- daily_states %>%
  mutate(yesterday_date = date - 1)

daily_states <- daily_states %>%
  left_join(daily_states, by = c("yesterday_date" = "date", "state" = "state")) %>%
  mutate(new_cases = cases.x - cases.y,
         new_deaths = deaths.x - deaths.y) %>% 
  select(date, state, fips.x, cases.x, deaths.x, new_cases, new_deaths) %>%
  replace(is.na(.),0) %>%
  rename(cases = cases.x,
         deaths = deaths.x,
         fips = fips.x) %>%
  left_join(state_pop, by = c("state" = "state")) %>%
  mutate(perc_pop_covid = round(100*cases / pop,3),
         perc_pop_death = round(100*deaths / pop,3),
         new_cases_prop = round(100*new_cases / pop,3),
         new_deaths_prop = round(100*new_deaths / pop,3)) %>%
  filter(date > '2020-03-01') %>%
  filter(!state %in% c("Northern Mariana Islands","Guam","Virgin Islands"))


####
# get state region data
state_regions <- read_csv(url("https://raw.githubusercontent.com/cphalpert/census-regions/master/us%20census%20bureau%20regions%20and%20divisions.csv")) %>%
  select(State,Region)

### make summary table
covid_summary <- daily_states %>%
  select(state, fips, cases, deaths, perc_pop_covid, perc_pop_death, pop) %>%
  summarize(total_cases = max(cases),
            total_deaths = max(deaths),
            perc_pop_covid = max(perc_pop_covid)/100,
            perc_pop_death = max(perc_pop_death)/100,
            pop = max(pop)) %>%
  left_join(state_regions, by = c("state" = "State")) %>%
  filter(!is.na(Region)) %>%
  arrange(perc_pop_covid) %>%
  mutate(state = factor(state, levels = state))
```
**Last updated:** *`r format(as.Date(Sys.Date(),"%Y-%m-%d"), "%B %d, %Y")`*      
Best viewed on desktop.  
  
# U.S. Covid-19 Vaccine Campaign Progress  
```{r web scrape vax, echo = FALSE, message=FALSE, warning=FALSE,fig.align='center'}
# scrape vax data from NY Times website https://www.nytimes.com/interactive/2020/us/covid-19-vaccine-doses.html
vax_html_raw <- content(GET('https://www.nytimes.com/interactive/2020/us/covid-19-vaccine-doses.html'))

# turn into manageable table
vax <- vax_html_raw %>%
  html_node('table') %>%
  html_table() 

full_states <- read_csv("full_states_nyt_html.csv", 
    col_names = FALSE)

names(vax) <- c(1,2,3,4,5,6)

vax <- vax %>%
  rename(state = 1) %>%
  rename(perc_vax = 2,
    doses_received = 4,
         shots_given = 5) %>%
 mutate(doses_received = parse_number(doses_received),
        shots_given = parse_number(shots_given),
        perc_vax = as.numeric(sub("%", "", perc_vax))/100) %>%
  select(1,2,4,5) 

for(i in 1:nrow(vax)){
  vax$state[i] <- full_states$X1[i]

}

# join with covid_summary table and add more summary stats
master_summary <- covid_summary %>%
  left_join(vax, by = "state") %>%
  mutate(perc_protection = perc_vax/2 + perc_pop_covid) %>%
  mutate(perc_herd_immunity = perc_protection/0.7) %>%
  mutate(perc_vax_given = round(100*shots_given/doses_received,1))

# import shapefile and fortify for ggplot (source: https://www.marmoe.xyz/2018/09/04/shapefiles-in-r-with-ggplot2-rgdal/)
map_shp <- broom::tidy(us_reshaped)
us_reshaped$id <- row.names(us_reshaped)
map_shp <- left_join(map_shp, us_reshaped@data)

# add covid and vax info
map_data <- left_join(map_shp, master_summary, by = c("NAME" = "state"))

# compile state centroids 
state_centers <- as.data.frame(rgeos::gCentroid(us_reshaped, byid=TRUE))
state_abbv <- map_data %>%
  select(STUSPS) %>%
  rename(abbv = STUSPS) %>%
  distinct() 
state_centers <- cbind(state_centers,state_abbv) %>%
  filter(!abbv == "DC")

# topline values for map titles
nyt_html_raw <- content(GET("https://www.nytimes.com/interactive/2020/us/covid-19-vaccine-doses.html")) %>%
  html_node('p.g-body') %>%
  html_text()

# first doses
cdc_vax_total_doses<- as.numeric(unlist(regmatches(nyt_html_raw,
                             gregexpr("[[:digit:]]+\\.*[[:digit:]]*",nyt_html_raw))))


cdc_vax_total_first_doses <- round(cdc_vax_total_doses[1],2)
cdc_vax_total_second_doses <- round(cdc_vax_total_doses[3],2)
# total doses 
cdc_vax_total_doses <- cdc_vax_total_first_doses + cdc_vax_total_second_doses       
       
# total first doses vax     
total_vax_perc <- format(round(100000000*cdc_vax_total_first_doses/sum(master_summary$pop),2), scientific = F)


# make map
ggplot(map_data, aes(x = long, 
                     y = lat, 
                     group = group, 
                     fill = perc_vax)) +
  
  geom_polygon(color = "white",
               size = .2) +
  
    geom_text(data=state_centers, aes(x, y, group = NULL, fill = NULL, label = abbv),
                    color = "black", size=2) +
  
  theme_map() +
  
  labs(x = NULL, 
       y = NULL,
       title = paste0("~", total_vax_perc,"% of U.S. population vaccinated against Covid-19"),
       subtitle = paste0("Approximately ", cdc_vax_total_first_doses," million Americans vaccinated since December 2020.\n",
                         "Last updated: ", format(as.Date(Sys.Date(),"%Y-%m-%d"), "%B %d, %Y")),
       caption = "Data from The New York Times; the CDC; U.S. Department of Health and Human Services; and state surveys and announcements.",
       fill = NULL) +
  
  scale_fill_gradient(low = "grey", high = "light blue", labels = scales::percent) +
  
  theme_map() +

  theme(legend.position = "right",
        plot.caption = element_text(face = "italic"),
        plot.title = element_text(size=16, face = "bold"),
        text=element_text(family="Georgia"))

ggsave(filename = "vax_map.png", device = "png", width = 7, height = 5)

``` 
  
### Daily First Doses Given in U.S.
```{r time series of vaccinations, echo = FALSE, message=FALSE, warning=FALSE,fig.align='center'}
# prep daily dose data by subtracting all past vax with CDC total from today
days_since_dec_14 <- as.numeric(Sys.Date() - as.Date('2020-12-14'))
daily_vax_doses <- read_csv("vax_time_series.csv") 
 
historic_vax_total <- sum(daily_vax_doses$daily_vaccinations)
today_vax_doses <- as.numeric(round(cdc_vax_total_doses - historic_vax_total,1))

new_daily_row <- data.frame(Sys.Date(),today_vax_doses)
names(new_daily_row) <- c("date","daily_vaccinations")

daily_vax_doses <- rbind(daily_vax_doses,new_daily_row)

# get average rate accounting for 2020 reporting buildup on Jan 1 2021
average_daily_rate <- round(sum(daily_vax_doses$daily_vaccinations)/days_since_dec_14, 2)

nrow<- nrow(daily_vax_doses)
seven_day_average_rate <- round(sum(daily_vax_doses$daily_vaccinations[(nrow-6):nrow])/7,2)

# overwrite existing csv to include today's number, to be used tomorrow
# write_csv(daily_vax_doses, "vax_time_series.csv")

# remove backlog for smooth function
daily_vax_doses_no_backlog <- daily_vax_doses %>%
  filter(date!='2021-01-01')

# ready now to plot
daily_vax_doses %>%
ggplot(aes(x = date, y = daily_vaccinations)) +
  
  geom_col(fill = "#65DDEB", 
           alpha = .6, 
           size = 1) + 
  
  geom_smooth(data = daily_vax_doses_no_backlog, 
              aes(x = date, y = daily_vaccinations), 
              color = "#659AEB", 
              size = 1.5, 
              se=FALSE,
              fullrange=FALSE) +
  
  annotate(geom="text", 
           x= daily_vax_doses$date[6], 
           y = 2.35, 
           label= "Includes backlog from Dec. 14 - Jan. 1, 2021", 
           color = "black",
           size = 3,
           font = "Georgia") +
  
  theme_tufte() +
  
  scale_x_date(date_breaks = "5 days",
              date_labels = ("%B %d")) +
  
  scale_y_continuous(n.breaks = 8) +
  
  labs(x = NULL,
       y = "Total doses (in millions)",
       title = paste0("~", cdc_vax_total_doses, " million total doses given in U.S."),
       subtitle = paste0("Since vaccinations began, the U.S. has averaged ", average_daily_rate, " million total doses per day.\n",
                         "Seven day rolling average: ", seven_day_average_rate, " million total doses per day.\n",
                         "Last updated: ", format(as.Date(Sys.Date(),"%Y-%m-%d"), "%B %d, %Y")),
       caption = "Data from CDC.gov") +
  
  theme(plot.caption = element_text(size = 12, face = "italic"),
        plot.title = element_text(size=16, face = "bold"),
        plot.subtitle = element_text(size=12),
        text=element_text(family="Georgia"),
        axis.title.y = element_text(size = 12),
        axis.text = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(angle = 40, vjust = 1, hjust = 1, size = 10),
        panel.grid.major.y = element_line(colour = "grey", size = .5),
        panel.grid.minor.y = element_line(colour = "grey90", size = 0.3))

ggsave(filename = "daily_vax.png", device = "png", width = 10, height = 6)

```   
  
  
The U.S. Covid-19 vaccination campaign began in earnest in December, 2020 after the dual emergency use authorizations of the [Pfizer-BioNTech](https://www.fda.gov/news-events/press-announcements/fda-takes-key-action-fight-against-covid-19-issuing-emergency-use-authorization-first-covid-19) and [Moderna](https://www.fda.gov/news-events/press-announcements/fda-takes-additional-action-fight-against-covid-19-issuing-emergency-use-authorization-second-covid#:~:text=The%20emergency%20use%20authorization%20allows,years%20of%20age%20and%20older.) mRNA vaccine candidates. The vaccines contain "a small piece of the Covid-19 virus’s mRNA that instructs cells in the body to make the virus’s distinctive 'spike' protein. After a person receives this vaccine, their body produces copies of the spike protein, which does not cause disease, but triggers the immune system to learn to react defensively, producing an immune response against SARS-CoV-2."  

  
### 100 Million Shots in 100 Days?  

```{r 100_days_projected graph, echo = FALSE, message=FALSE, warning=FALSE,fig.align='center'}
projected_vax_100 <- daily_vax_doses

current_vax_total <- sum(projected_vax_100$daily_vaccinations)
date <- Sys.Date() + 1

while(current_vax_total <= 100) {
  
  add_row <- data.frame(date,seven_day_average_rate)
  
  names(add_row) <- c("date","daily_vaccinations")

  projected_vax_100 <- rbind(projected_vax_100,add_row)
  
  current_vax_total <- current_vax_total + seven_day_average_rate
  date <- date + 1
}

for(i in 1:nrow(projected_vax_100)){
  projected_vax_100$total[i] <- sum(projected_vax_100$daily_vaccinations[1:i])
}

first_area <- projected_vax_100 %>% filter(date <= Sys.Date())
second_area <- projected_vax_100 %>% filter(date > max(first_area$date)-1)

projected_vax_100 %>%
ggplot(aes(x = date, y = total)) +
  
  geom_point(fill = "#65DDEB", 
           alpha = 1, 
           size = .5) + 
  
  geom_line() +
  
  geom_area(data = first_area,
            alpha = .4,
            aes(fill = "Doses given")) +
  
    geom_area(data = second_area,
            alpha = .4,
            aes(fill = "Projected doses")) +

  
  # 100 mil doses day
  geom_col(aes(x = projected_vax_100$date[nrow(projected_vax_100)],
               y = 1.13,
               fill = "100 million doses")) +
  
    # biden 100 days
  geom_col(aes(x = date('2021-04-30'),
           y = 1.13,
           fill = "Biden's 100 days")) +
  
  # geom_smooth(data = daily_vax_doses_no_backlog, 
  #             aes(x = date, y = daily_vaccinations), 
  #             color = "#659AEB", 
  #             size = 1.5, 
  #             se=FALSE,
  #             fullrange=FALSE) +
  
  
  theme_tufte() +
  
  scale_x_date(date_breaks = "10 days",
              date_labels = ("%B %d")) +
  
  scale_y_continuous(n.breaks = 8) +
  
  scale_fill_manual(values=c("#10B626","red", "#648D69", "grey")) +
  
  labs(x = NULL,
       y = "Total doses (in millions)",
       fill = NULL,
       title = paste0("Biden's promise: 100 million shots in 100 days"),
       subtitle = paste0("Projected 100 million doses date: ", format(projected_vax_100$date[nrow(projected_vax_100)],"%B %d, %Y"), "\n",
                         "7 day rolling average: ", seven_day_average_rate, " million total doses per day"),
       caption = paste0("Last updated: ", format(as.Date(Sys.Date(),"%Y-%m-%d"), "%B %d, %Y"))) +
  
  theme(plot.caption = element_text(size = 12, face = "italic"),
        plot.title = element_text(size=16, face = "bold"),
        plot.subtitle = element_text(size=12),
        text=element_text(family="Georgia"),
        axis.title.y = element_text(size = 12),
        axis.text = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(angle = 40, vjust = 1, hjust = 1, size = 10),
        panel.grid.major.y = element_line(colour = "grey", size = .5),
        panel.grid.minor.y = element_line(colour = "grey90", size = 0.3))

ggsave(filename = "100_days_projected.png", device = "png", width = 10, height = 6)

```
  
  
***
# Covid-19 Vaccine Distribution
### U.S. Allocation vs. Distribution
```{r bar plot distribution, echo = FALSE, message=FALSE, warning=FALSE,fig.align='center'}
# compare vaccine allocated, distributed, and given across states
simple_bar <- vax %>%
  filter(state == "U.S. Total*") %>%
  pivot_longer(cols = c("doses_received","shots_given"), names_to = "metric") %>%
  select(metric,value) %>%
  mutate(perc = scales::percent(value/max(value),2))

simple_bar$metric[1] <- "Vaccines distributed to states"
simple_bar$metric[2] <- "Total doses given"
  


ggplot(simple_bar, aes(x= rev(metric), 
                               y = value, 
                               fill = factor(metric, levels = unique(metric)))) +
  
  geom_col(position = "dodge") +
  
  geom_label(
    label=simple_bar$perc[2],
    x=simple_bar$metric[1],
    y=simple_bar$value[2] + simple_bar$value[2]/5,
    show.legend = FALSE,
    fill = "white",
    size = 8
  ) +
  
  geom_segment(aes(x = 1.5,
                   y = simple_bar$value[1],
                   xend = 1.9,
                   yend = simple_bar$value[2] + 2*simple_bar$value[2]/5),
               arrow = arrow(type = "closed")) +

  
  labs(x = NULL,
       y = NULL,
       caption = paste0("Data from CDC.gov. Last updated: ", format(as.Date(Sys.Date(),"%Y-%m-%d"), "%B %d, %Y")),
       title = NULL,
       fill = NULL) +
    
  scale_y_continuous(labels = scales::comma) +
  
  theme_tufte() +
  
  theme(plot.caption = element_text(size = 12, face = "italic"),
        plot.title = element_text(size=16, face = "bold"),
        plot.subtitle = element_text(size=12),
        text=element_text(family="Georgia"),
        axis.title.y = element_text(size = 12),
        axis.text = element_text(size = 14, face = "bold"),
        axis.text.x = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major.y = element_line(colour = "grey", size = .5),
        panel.grid.minor.y = element_line(colour = "grey90", size = 0.3),
        legend.position = "top",
        legend.text = element_text(size = 10))

ggsave(filename = "total_vax.png", device = "png", width = 7, height = 5)

``` 
  
### Flow of Vaccine Distribution
```{r vax flow, echo = FALSE, message=FALSE, warning=FALSE,fig.align='center', out.width="100%"}
# build nodes
flow_nodes <- data.frame(c(1:4),c("Operation Warp Speed", "Vaccine distributed to states", "First doses administered", "Second doses administered"))
names(flow_nodes) <- c("id","label")

# build edges
flow_edges <- data.frame(c("Operation Warp Speed", "Vaccine distributed to states", "First doses administered"), 
                         c("Vaccine distributed to states", "First doses administered", "Second doses administered"),
                         c(simple_bar$value[1], cdc_vax_total_first_doses*10^6, cdc_vax_total_second_doses*10^6))
names(flow_edges) <- c("source","destination","weight")
flow_edges <- flow_edges %>%
  mutate(weight = round(weight,0))

# prep for sankey
flow_edges <- flow_edges %>%
  left_join(flow_nodes, by = c("source" = "label")) %>% 
  rename(from = id)

flow_edges <- flow_edges %>% 
  left_join(flow_nodes, by = c("destination" = "label")) %>% 
  rename(to = id)

flow_edges <- select(flow_edges, from, to, weight)

# manipulate nodes and edges for d3 specification
flow_nodes_d3 <- mutate(flow_nodes, id = id - 1)
flow_edges_d3 <- mutate(flow_edges, from = from - 1, to = to - 1)

# prepare colour scale
sankey_color <-'d3.scaleOrdinal() .range(["#3395FF","#EBE965","#AAEB65", "green"])'

sankeyNetwork(Links = flow_edges_d3, Nodes = flow_nodes_d3, Source = "from", Target = "to",
              NodeID = "label", Value = "weight", fontSize = 16,
              sinksRight = F,
              iterations = 50,
              nodeWidth = 30,
              nodePadding = 30,
              colourScale = sankey_color)
```


### State Allocation vs. Distribution
```{r state bar plot distribution, echo = FALSE, message=FALSE, warning=FALSE,fig.align='center', fig.height= 10, out.width= "100%"}
# compare vaccine allocated, distributed, and given across states
bar_stats <- master_summary %>%
  select(state, doses_received, shots_given) %>%
  rename('Doses allocated to state' = doses_received,
         'Total doses given' = shots_given) %>%
  mutate(state = factor(state, levels = state)) %>%
  pivot_longer(!state, "metric", "value") 
  
bar_stats <- master_summary %>%
  select(state,Region, perc_vax_given) %>%
  right_join(bar_stats, by = "state") %>%
  arrange(value) %>%
  mutate(state = factor(state, levels = unique(state))) %>%
  filter(!state == "District of Columbia") 

ggplotly(ggplot(bar_stats, aes(x= state, 
                               y = value, 
                               fill = factor(metric, levels = unique(metric)),
                               text = paste0("<b>", state, "</b> <br>",
                                             "<b>% received shots given: </b> <br>", 
                                             perc_vax_given, "% <br>",
                                             "<br>",
                                                 "<b>", metric,":</b><br>",
                                                 value))) +
  
  geom_col(position = "dodge") +
  
  facet_wrap(~Region, ncol = 2, scales = "free_y") +
  
  labs(x = NULL,
       y = NULL,
       caption = "Data from The New York Times.",
       fill = NULL,
       title = NULL) +
    
  scale_y_continuous(labels = scales::comma) +
  
  coord_flip() +
  
  theme_tufte() +
  
  theme(axis.text.y = element_text(size=12),
        text=element_text(family="Georgia"),
        plot.caption = element_text(face = "italic"),
        plot.title = element_text(size=18, face = "bold"),
        legend.position = "bottom",
        strip.text.x = element_text(size = 14, face = "bold"),
        panel.grid.major.x = element_line(colour = "grey", size = 0.5),
        axis.text.x = element_text(face = "bold", angle = 30)),
  
tooltip = "text") %>%

  #ggplotly options
  plotly::config(displayModeBar = FALSE) %>%
  layout(hoverlabel = list(bgcolor = "white",
                           font = list(family = "Georgia")),
         legend = list(x=.3,
                       y = 1.1,
                       orientation = 'h',
                       itemsizing='constant'))

``` 
  
  
Though the vaccines were [developed and approved in record time](https://learningenglish.voanews.com/a/how-covid-19-vaccines-developed-in-record-time/5703115.html), the vaccine distribution campaign is off to a [slow and concerning start](https://www.nbcnews.com/politics/white-house/biden-talk-covid-harris-gets-vaccine-n1252464). Though the federal government's [Operation Warp Speed](https://www.hhs.gov/coronavirus/explaining-operation-warp-speed/index.html) fell far short of its [initial goal to deliver 20 million vaccinations](https://www.npr.org/sections/coronavirus-live-updates/2020/12/31/952208601/u-s-likely-will-miss-goal-of-vaccinating-20-million-by-the-new-year) by the end of December, improvements in distribution make it quite clear the U.S. will hit the Biden administration's goal of [100 million vaccine doses in his first 100 days](https://www.theguardian.com/us-news/2020/dec/08/joe-biden-coronavirus-vaccine-health-team). It now appears Biden is aiming for 1.5 million doses per day in his 100 days, which again, seems reasonable if not certain.  

But forget goals - what if the U.S. distributed all the vaccine supply promised to it by Moderna and Pfizer (220 million doses by end of Q1 2021)? We could hit herd immunity be early June.  
  
***
# When Will the U.S. Achieve Herd Immunity?  
```{r herd immunity map, echo = FALSE, message=FALSE, warning=FALSE,fig.align='center'}


# days since dec 14 / % percent to herd * 100 = total days
# herd immunity date = total days + dec 14

days_since_dec_14 <- as.numeric(Sys.Date() - as.Date('2020-12-14'))

total_days <- round(days_since_dec_14 * 70 / as.numeric(total_vax_perc),0)

vax_day <- format(as.Date('2020-12-14') + total_days, "%B %d, %Y")

# new herd immmunity projection

# seven day covid case average
seven_day_average_covid_case_rate <- daily_states %>%
  filter(date > Sys.Date() - 8) %>%
  group_by(date) %>%
  summarize(seven_day_state_cases = sum(new_cases)) %>%
  ungroup() %>%
  summarize(seven_day_mean_cases = round(mean(seven_day_state_cases),0)/1000000)

seven_day_average_covid_case_rate <- seven_day_average_covid_case_rate$seven_day_mean_cases[1]

# full projected
full_herd_projection <- daily_vax_doses %>%
  rename(protection_mil = 2) %>%
  mutate(protection_mil = protection_mil/2)

full_herd_projection$protection_mil[1] <- full_herd_projection$protection_mil[1] + sum(master_summary$total_cases)/1000000

current_herd_total <- sum(full_herd_projection$protection_mil)

date <- Sys.Date() + 1

while(current_herd_total <= 231) {
  
  add_row <- data.frame(date,seven_day_average_rate/2 + seven_day_average_covid_case_rate)
  
  names(add_row) <- c("date","protection_mil")

  full_herd_projection <- rbind(full_herd_projection,add_row)
  
  current_herd_total <- current_herd_total + seven_day_average_rate/2 + seven_day_average_covid_case_rate
  date <- date + 1
}

for(i in 1:nrow(full_herd_projection)){
  full_herd_projection$total[i] <- sum(full_herd_projection$protection_mil[1:i])
}

# topline % to immunity stat
# total covid cases + total shots given / total us population / .7
# us_perc_to_immunity <- round(100*(sum(master_summary$total_cases) + cdc_vax_total_doses*1000000) / 
#   sum(master_summary$pop) / 
#   .7,1)

us_perc_to_immunity <- full_herd_projection %>% filter(date == Sys.Date())
us_perc_to_immunity <- round(100*us_perc_to_immunity$total[1]/231,1)


# make map
ggplot(map_data, aes(x = long, 
                     y = lat, 
                     group = group, 
                     fill = perc_herd_immunity)) +
  
  geom_polygon(color = "white",
               size = .2) +
  
  geom_text(data=state_centers, aes(x, y, group = NULL, fill = NULL, label = abbv),
                    color = "black", size=2) +
  
  theme_map() +
  
  labs(x = NULL, 
       y = NULL,
       title = paste0("~", us_perc_to_immunity, "% of the way to 70% U.S. herd immunity threshold"),
       subtitle = paste0("Projected herd immunity date: ",format(full_herd_projection$date[nrow(full_herd_projection)],"%B %d, %Y"),"\nLast updated: ",
         format(as.Date(Sys.Date(),"%Y-%m-%d"), "%B %d, %Y")),
       caption = "Data from The New York Times and the CDC.",
       fill = "% to herd immunity") +
  
  scale_fill_gradient(low = "grey", high = "PaleGreen", labels = scales::percent) +
  
  theme_map() +
  
  theme(legend.position = "right",
        plot.caption = element_text(face = "italic"),
        plot.title = element_text(size=16, face = "bold"),
        text=element_text(family="Georgia"))

ggsave(filename = "herd_map.png", device = "png", width = 7, height = 5)

# deaths to herd immunity
deaths_to_herd <- format(round(sum(master_summary$total_deaths)/(sum(master_summary$perc_pop_covid)/100) - sum(master_summary$total_deaths),0),format = "d", big.mark = ",")

# herd date
herd_date <- format(full_herd_projection$date[nrow(full_herd_projection)],"%B %d, %Y")
```
  
  
```{r projected_herd_timeline, echo = FALSE, message=FALSE, warning=FALSE,fig.align='center'}
herd_first_area <- full_herd_projection %>% filter(date <= Sys.Date())
herd_second_area <- full_herd_projection %>% filter(date > max(herd_first_area$date))

full_herd_projection %>%
ggplot(aes(x = date, y = total)) +
  
  geom_point(fill = "#65DDEB", 
           alpha = 1, 
           size = .1) + 
  
  geom_line() +
  
  geom_area(data = herd_first_area,
            alpha = .4,
            aes(fill = "People protected")) +
  
    geom_area(data = herd_second_area,
            alpha = .4,
            aes(fill = "Projected people protected")) +
  
    # herd day
  geom_col(aes(x = max(full_herd_projection$date),
           y = .83,
           fill = "Herd Immunity Day!!")) +
  
  theme_tufte() +
  
  scale_x_date(date_breaks = "25 days",
              date_labels = ("%B %d"),
              limits = c(min(full_herd_projection$date), max(full_herd_projection$date)+1)) +
  
  scale_y_continuous(n.breaks = 5) +
  
  scale_fill_manual(values=c("#10B626","yellow", "grey")) +
  
  labs(x = NULL,
       y = "People protected (in millions)",
       fill = NULL,
       title = paste0("When will we reach herd immunity?"),
       subtitle = paste0("At 70% threshold, projected date: ", format(full_herd_projection$date[nrow(full_herd_projection)],"%B %d, %Y"), "\n"),
       caption = paste0("Last updated: ", format(as.Date(Sys.Date(),"%Y-%m-%d"), "%B %d, %Y"))) +
  
  theme(plot.caption = element_text(size = 12, face = "italic"),
        plot.title = element_text(size=16, face = "bold"),
        plot.subtitle = element_text(size=12),
        text=element_text(family="Georgia"),
        axis.title.y = element_text(size = 12),
        axis.text = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(angle = 40, vjust = 1, hjust = 1, size = 10),
        panel.grid.major.y = element_line(colour = "grey", size = .5),
        panel.grid.minor.y = element_line(colour = "grey90", size = 0.3))

ggsave(filename = "projected_herd_date.png", device = "png", width = 10, height = 6)

```
  
  
The majority of public health experts have estimated general herd immunity is achieved somewhere around [70% protection](https://www.mayoclinic.org/diseases-conditions/coronavirus/in-depth/herd-immunity-and-coronavirus/art-20486808), though some experts have recently estimated that [higher protection rates](https://www.reuters.com/article/health-coronavirus-usa/fauci-says-herd-immunity-could-require-nearly-90-to-get-coronavirus-vaccine-idINL1N2J411V) are necessary. 

The percent to herd immunity calculations presented in the figure above assume this 70% protection rate and some form of [protection against reinfection](https://www.cancer.gov/news-events/cancer-currents-blog/2020/coronavirus-antibodies-protect-against-future-infection#:~:text=Using%20real%2Dworld%20data%20from,being%20reinfected%20with%20the%20virus%20.) from Covid-19 antibodies. While the true herd immunity protection threshold may be higher than 70%, the true number of U.S. Covid-19 cases may also be [substantially underestimated](https://publichealth.berkeley.edu/news-media/research-highlights/us-covid-19-cases-may-be-substantially-underestimated/). 

$\text{herd immunity date} = D$

$$\text{ where:}$$

$$\sum_{n=a}^{i} C_n +  \sum_{n=b}^{i} \frac{V_n}{2}+ \sum_{n=i}^{D} (\frac{\hat V_n}{2} + \hat C_n) = .70*USpop$$
$$\text{and:}$$

$a = \text{date of first Covid-19 case}$

$i = \text{today's date}$

$C = \text{daily Covid-19 cases confirmed}$

$b = \text{date of first vaccine dose given}$

$V = \text{daily vax doses given}$

$\bar v = \text{seven day rolling average of daily vax doses given}$

$\hat V = \text{projected daily vax doses given} = \bar v$

$\bar c = \text{seven day rolling average of daily Covid cases confirmed}$

$\hat C = \text{projected daily Covid-19 cases} = \bar c$

$USpop = 330,000,000$





Such uncertainty could lead to herd immunity being achieved either more quickly or more slowly than projected above. Thus, these figures are meant to illustrate a rough approximation of the country's progress towards herd immunity, rather than a definitive projection.   

**At the rolling seven-day average vaccination and case rates, the U.S. is projected to achieve herd immunity `r herd_date`.** All assumptions considered, this is likely a conservative date - assuming single-dose vaccines are introduced into distribution from Johnson & Johnson and/or AstraZeneca, and the rate of daily vaccinations continues to increase.   

What is more certain is that to achieve herd immunity by Covid-19 infection alone, as some [fringe public health commentators](https://www.politico.com/news/2020/12/16/trump-appointee-demanded-herd-immunity-strategy-446408) and [right-wing politicians](https://www.washingtonpost.com/politics/trump-coronavirus-scott-atlas-herd-immunity/2020/08/30/925e68fe-e93b-11ea-970a-64c73a1c2392_story.html) have previously suggested the U.S. do, **approximately `r deaths_to_herd` more Americans would die!**  

```{r reactable_summary, echo = FALSE, message=FALSE, warning=FALSE,fig.align='center'}

vax_table <- master_summary %>%
  filter(state != "District of Columbia") %>%
  arrange(state) %>%
  select(1,9, 10, 13, 8, 12) %>%
  mutate(perc_herd_immunity = round(perc_herd_immunity,2)) %>%
  mutate(perc_vax = scales::percent(perc_vax, accuracy = .1),
         perc_vax_given = scales::percent(perc_vax_given/100, accuracy = .1),
         perc_herd_immunity = scales::percent(perc_herd_immunity, accuracy = 1)) %>%
  rename(State = state,
         `Doses Received` = doses_received,
         `Shots Givens` = shots_given,
         `% Vaccinated` = perc_vax,
         `% Doses Used` = perc_vax_given,
         `% to Herd Immunity` = perc_herd_immunity) 
  
reactable(vax_table,
          defaultSorted = list("State" = "asc"),
          searchable = TRUE,
          bordered = TRUE,
          striped = TRUE,
          highlight = TRUE,
          compact = TRUE, 
          resizable = TRUE)

```
  
***
# Covid-19 Case Data 
### Population With Confirmed Covid-19 Diagnosis To Date
```{r bar charts, echo = FALSE, warning = FALSE, message = FALSE, fig.align='center', fig.height= 14, out.width= "100%"}

ggplotly(ggplot(covid_summary, aes(x = state, 
                                   y = perc_pop_covid, 
                                   color = Region, 
                                   # for plotly display box later on
                                   text = paste0("<b>",state, "</b> <br>",
                                                 "<b>% of pop. diagnosed: </b> <br>",
                                                 round(perc_pop_covid*100,2),"% <br>",
                                                 "<br>",
                                                 "<b>total # of cases: </b><br>", 
                                                 total_cases))) +
  
  geom_segment(aes(xend = state, yend=0)) +
  
  geom_point(size = 3, color = "black") +
  
  scale_y_continuous(labels= scales::percent) +
  
  coord_flip() +
  
  labs(x = NULL,
       y = NULL,
       caption = "Data from The New York Times and U.S. Census Bureau.",
       title= NULL) +
  
  facet_wrap(~Region, ncol = 1, scales = "free_y") +
  
  theme_tufte() +
  
  theme(axis.text.y = element_text(size=12),
        text=element_text(family="Georgia"),
        plot.caption = element_text(face = "italic"),
        plot.title = element_text(size=16, face = "bold"),
        legend.position = "none",
        strip.text.x = element_text(size = 14, face = "bold"),
        panel.grid.major.x = element_line(colour = "grey", size = 0.5),
        axis.text.x = element_text(face = "bold")),

  tooltip = "text") %>%
  
  #ggplotly options
  plotly::config(displayModeBar = FALSE) %>%
  layout(hoverlabel = list(bgcolor = "white",
                           font = list(family = "Georgia")))
```
  
***
### Time Series of Daily Covid-19 Cases Across U.S. States
```{r ridgeplot cases, echo=FALSE, fig.height = 20, fig.align='center'}
# summary stats for total cases
total_covid_cases <- format(sum(master_summary$total_cases), format = "d", big.mark = ",")

## ridge plot of daily new cases per state             
daily_states %>% 
  mutate(cum_cases = max(cases)) %>%
  arrange(desc(cum_cases)) %>%
  ungroup() %>%
  mutate(state = fct_reorder(state, cum_cases, max)) %>%
  
ggplot(aes(x = date, 
           y = state, 
           ordered = TRUE, 
           height = new_cases,
           fill = state)) +
  
  geom_density_ridges2(stat = "identity", 
                      alpha = .6, 
                      scale = 10,
                      size = .2) +
  
  theme_ridges() +
  
  labs(x = NULL, 
       y = NULL,
       title = paste0("New daily cases of Covid-19: ", total_covid_cases, " total"),
       caption = "Data from The New York Times and U.S. Census Bureau.") +
  
  scale_fill_cyclical(values = c("red","orange","yellow")) +
  
  theme(axis.text.y = element_text(size=12),
        axis.text.x = element_text(face = "bold"),
        panel.grid.major.x = element_blank(),
        text=element_text(family="Georgia"),
        plot.title = element_text(size=16, face = "bold", hjust = 0),
        plot.margin = margin(2,6,2,1,"mm"),
        plot.caption = element_text(face = "italic", size = 8)) 
  
#  coord_fixed(ratio = 20)
  
  
```
  
***
### Time Series of Daily Covid-19 Cases Proportional to State Populations
```{r ridgeplot proportion cases, echo=FALSE, fig.height = 18.5, fig.align='center'}
## ridge plot of daily new cases per state             
daily_states %>%
  mutate(prop_rank = sum(new_cases_prop)) %>%
  arrange(desc(prop_rank)) %>%
  ungroup() %>%
  mutate(state = fct_reorder(state, prop_rank, max)) %>%

ggplot(aes(x = date, 
           y = factor(state, levels = rev(unique(state))), 
           ordered = TRUE, 
           height = new_cases_prop,
           fill = state)) +
  
  geom_density_ridges2(stat = "identity", 
                      alpha = .6, 
                      scale = 6,
                      size = .2) +
  
  theme_ridges() +
  
  labs(x = NULL, 
       y = NULL,
       title = "New daily cases of Covid-19 -",
       subtitle = "as proportion of each state's population",
       caption = "Data from The New York Times and U.S. Census Bureau.") +
  
  scale_fill_cyclical(values = c("purple","blue","dark blue")) +
  
  theme(axis.text.y = element_text(size=12),
        axis.text.x = element_text(face = "bold"),
        panel.grid.major.x = element_blank(),
        text=element_text(family="Georgia"),
        plot.title = element_text(size=16, face = "bold"),
        plot.margin = margin(2,6,2,1,"mm"),
        plot.caption = element_text(face = "italic", size = 8)) 
  
#  coord_fixed(ratio = 20)
  
```

```{r mask wearing, echo = FALSE, include=FALSE}
# library(covidcast)
# 
# cc_data <- suppressMessages(
#   covidcast_signal(data_source = "fb-survey", signal = "smoothed_wearing_mask",
#                    start_day = "2020-04-06", end_day = "2020-12-18",
#                    geo_type = "state"))
# 
# st_crosswalk <- tibble(state = state.name) %>%
#   bind_cols(tibble(abb = state.abb)) %>% 
#   bind_rows(tibble(state = "District of Columbia", abb = "DC")) %>%
#   mutate(abb = str_to_lower(abb))
# 
# cc_data <- left_join(cc_data, st_crosswalk, by = c("geo_value" = "abb")) %>%
#   group_by(state) %>%
#   summarize(mean_mask_perc = round(mean(value),0))
#   
# prop_deaths <- daily_states %>%
#   summarize(cum_deaths = max(deaths),
#             pop = pop)
# 
# mask <- cc_data %>%
#   left_join(prop_deaths, by = c("state" = "state")) %>%
#   distinct() %>%
#   mutate(prop_deaths = round(100*cum_deaths/pop,3)) 
#   
# ggplot(mask,aes(x=prop_deaths, y = mean_mask_perc)) +
#   geom_point() +
#   geom_smooth(method=lm, se= FALSE)
#  scale_y_log10()
```
  

**Questions or comments?** *Please reach out!*  

  
**Data Sources:**  
- [CDC vaccine distribution data](https://covid.cdc.gov/covid-data-tracker/#vaccinations)  
- [NY Times Covid-19 github repo](https://github.com/nytimes/covid-19-data)  
  

*Created in R.*

```{r twitter, include=FALSE}
library(twitteR)

# connect to app
twitter_app_name <- "US_covid_vax"
twitter_api_key <- "STBer0BAcSL5LFnZ625Ek2nOF"
twitter_api_secret <- "u2boBBvcPFUnWbluC9YCxdCSTvq3F5szyJBkqjMW2XF1PTjYhJ"
twitter_bearer_token <- "AAAAAAAAAAAAAAAAAAAAAIUMLQEAAAAAe0G%2FxEMgbhwZ1ve99ZSq8Vs5SbY%3DAJBfaRfnIGa0MlcJZapmweod5wnEBrbgHybtyE7PDWBkNRyCFN"

setup_twitter_oauth(twitter_api_key, twitter_api_secret, "1345057778822504448-exYE3x4CTA33PabgXpQFUEyHYWwU59", "GUGstuQfV1GpPXwysI5snwJh3gon6Va6LKghAr8XNe5We")

# post a tweet from R
daily_tweet <- paste0("Daily update ", format(as.Date(Sys.Date(),"%Y-%m-%d"), "%m/%d/%Y"), ":\n",
                      "\n",
                      total_vax_perc, "% of the U.S. population has been vaccinated against Covid-19. That's ", cdc_vax_total_first_doses, " million first doses.\n",
                      "\n",
                      "The seven day rolling average is ", seven_day_average_rate, " million doses per day.\n",
                      "\n",
                      "The U.S. is ~", us_perc_to_immunity, "% of the way to herd immunity.\n",
                      "\n",
                      "Live tracker: https://bit.ly/covid_vax_tracker"
                      )

# tweet! and change image each day
tweet_image_func <- function(){
  x <- sample(1:4)
  if(x==1){
    "herd_map.png"
  } else if (x==2) {
    "vax_map.png"
  } else if (x==3){
    "daily_vax.png"
  } else {
    "projected_herd_date.png"
  }
}
tweet_image <- tweet_image_func()

# SECTION TO UNCOMMENT ONCE WHEN KNITTING
# updateStatus(text = daily_tweet,
#              mediaPath = tweet_image,
#              bypassCharLimit = TRUE)

# # overwrite existing csv to include today's number, to be used tomorrow
 # write_csv(daily_vax_doses, "vax_time_series.csv")

```
 