---
title: "The World Gets Better Every Year"
author: "Tanner Gildea"
date: "2019-03--5"
output: html_document
---

Scroll through your Twitter feed or flip to any news station, and you're likely to hear some news that leads you to believe the state of world affairs is close to collapse. 

These stories are true and reflect the immense suffering of specific people, in specific times and places. But what's lost amidst these stories is the larger narrative of progress humanity as a whole continues to make against the scourges of disease, poverty, and death. 

The underlying data is clear: the world as a whole gets better every year.

Harvard Professor of Psychology Steven Pinker's recent book ["Enlightenment Now: The Case for Reason, Science, Humanism, and Progress"](https://stevenpinker.com/publications/enlightenment-now-case-reason-science-humanism-and-progress) explores this progess in depth, drawing upon metrics of "life, health, prosperity, safety, peace, knowledge, and happiness" to make his case. 

Arguably, two of the most important measurements of human welfare are life expectancy and GDP per capita. Using data from the [Gapminder](https://www.gapminder.org) dataset, I've created a visualization that previews Pinker's broader argument. 




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(cache = TRUE)
library(tidyverse)
library(gapminder)
library(ggplot2)
library(gganimate)
```

```{r visualize}
gapminder %>% 
  mutate(continent = fct_collapse(continent,
                                  Asia =  c("Asia", "Oceania"))) %>% 
  
  filter(continent %in% c("Americas", "Africa", "Asia")) %>% 
  
  # changes order of continents in columns
  mutate(continent = fct_relevel(continent, c("Americas", "Africa", "Asia"))) %>% 
  
  ggplot(aes(x = gdpPercap, y = lifeExp, size = pop, color = country)) + 
  geom_point(alpha = .7, show.legend = FALSE) + 
  
  facet_wrap(~continent) + 
  
  scale_colour_manual(values = country_colors) +
  
  # makes x-axis scale more accessible
  scale_x_log10(breaks = c(1000,10000), labels = c("$1,000", "$10,000")) +
  
  # using {} within title allows variable year to change
  labs(title = "The World Gets Better Every Year: ",
       subtitle = "{frame_time}",
       caption = "Source: Gapminder Package",
       x = "GDP Per Capita",
       y = "Life Expectancy") +
  
  # gganimate part
  transition_time(year) +
  ease_aes('linear')
```